---
title: "intense rainfall events in R"
author: "Tyler Mitchell"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "stylesheet.css"
---

```{css, echo=FALSE}
.theme {
  background-color: #f7f4f9;
  border: 2px black;
  font-weight: bold;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(class.source = "theme")
```

## Introduction

I am interested in tree radial growth responses to intense precipitation. Here, we define intense precipitation as any daily precipitation value that is \>2.0 standard deviations above the long-term mean [1]. This classification technique is related to other measures of precipitation intensity, which is an area of continued research interest [e.g., 2, 3, 4].

A member of my doctoral committee suggested that it could be helpful to share a standardized method to calculate intense rain events (IREs), so here we are!

## Data inputs

A daily summary weather station .csv file from NOAA's Climate Data Online portal [5], preferably with a high percentage of data coverage.

## Process

```{r message = FALSE}
require("tidyverse")
```

```{r}
# read climate data ----
data <- read.csv("2917741.csv")
head(data)
```

```{r}
# convert NOAA's DATE into three separate variables using tidyr ----
data <- tidyr::separate(data, DATE, sep="-", into = c("year", "month", "day"))
data$year <- as.numeric(data$year)
data$month <- as.numeric(data$month)
data$day <- as.numeric(data$day)

# remove missing precipitation observations ----
data <- data[!is.na(data$PRCP),]

# subset months of interest for intense rainfall calculation, here: july (7) ----
selected.month <- 7
data <- data[data$month == selected.month,]

# standard-normal (z) scores for precipitation and selection of intensity threshold ----
data$z <- scale(data$PRCP)
threshold <- 2.00
data$ire <- ifelse(data$z > threshold, 1, 0)

# retain precipitation amounts from intense rainfall event (ire) selection ----
data$ireppt <- ifelse(data$ire == 1, data$PRCP, 0)

# summarize intense rainfall events by year ----
ire.frequency <- setNames(aggregate(data$ire ~ data$year, FUN = sum),
                   c("year", "ire.frequency"))

ire.precipitation <- setNames(aggregate(data$ireppt ~ data$year, FUN = sum),
                              c("year", "ire.precipitation"))

ire <- merge(ire.frequency, ire.precipitation, by = "year")
ire
```

```{r}
# another workflow using the lubridate package and pipes to calculate all years 
# and months at once. calculates standard normal (z) scores based on ALL months distribution ----
require("lubridate")
require("tidyverse")
data <- read.csv("2917741.csv")

# remove missing precipitation observations ----
data <- data[!is.na(data$PRCP),]

# standard-normal (z) scores for precipitation and selection of intensity threshold ----
data$z <- scale(data$PRCP)
threshold <- 2.00
data$ire <- ifelse(data$z > threshold, 1, 0)

# retain precipitation amounts from intense rainfall event (ire) selection ----
data$ireppt <- ifelse(data$ire == 1, data$PRCP, 0)

data <- data %>% 
  group_by(year = year(DATE), month = month(DATE)) %>% 
  summarise_if(is.numeric, sum)

data[c(1:2,6,8:9)]
```

## References

[1] Mitchell, T. J., Knapp, P. A., & Patterson, T. W. (2020). [The importance of infrequent, high-intensity rainfall events for longleaf pine (Pinus palustris Mill.) radial growth and implications for dendroclimatic research.](https://doi.org/10.1016/j.tfp.2020.100009) Trees, Forests and People, 1, 100009.

[2] Post, A. K., & Knapp, A. K. (2020). [The importance of extreme rainfall events and their timing in a semi‐arid grassland.](https://doi.org/10.1111/1365-2745.13478) Journal of Ecology, 108(6), 2431-2443.

[3] Griffin‐Nolan, R. J., Slette, I. J., & Knapp, A. K. (2021). [Deconstructing precipitation variability: Rainfall event size and timing uniquely alter ecosystem dynamics.](https://doi.org/10.1111/1365-2745.13724) Journal of Ecology, 109(9), 3356-3369.

[4] Moustakis, Y., Papalexiou, S. M., Onof, C. J., & Paschalis, A. (2021). [Seasonality, intensity, and duration of rainfall extremes change in a warmer climate.](https://doi.org/10.1029/2020EF001824) Earth's Future, 9(3), e2020EF001824.

[5] <https://www.ncdc.noaa.gov/cdo-web/results>
